---
title: "AD_Spatial_Transcriptome"
subtitle: "Script to make Pseudo-bulk matrices"
author:  |
 | Rush Alzheimer’s Disease Center
 | Chicago, IL 60612
date: "`r Sys.Date()`"
output: 
 rmarkdown::html_document:
   theme: united
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true  
---

```{r clean.variables, echo=FALSE}
#This command clean all variables. BE CAREFULL!!! 
rm(list = setdiff(ls(), lsf.str()))
```

```{r load.packages, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
library(Seurat)
library(tidyverse)
library(kableExtra)
library(furrr)
library(future)
future::plan(multicore)

knitr::opts_chunk$set( 
  warning=FALSE,
  message=FALSE)
```

```{r Helper Functions, echo=FALSE}
createDT <- function(DF, caption="", scrollY=500){
  data <- DT::datatable(DF, caption=caption,
    extensions =  'Buttons',
    options = list( dom = 'Bfrtip', 
                    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                    scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,  
                      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    )
  ) 
   return(data)
}

make_pseudo_matrices <- function(metadata.filt, 
                                 sample_ids,
                                 outputRdata = paste0(work_dir,"pseudo_bulk/by_donor_section/",output_prefix,"_pseudo_counts.RData"),
                                 nspots = 50){
  read_by_donor <- function(sample){
    #sample = "B16002"
    cnt_file = data.df[data.df$sample_id == sample,]$file
    suppressWarnings(dat <- data.table::fread(cnt_file) %>% as.data.frame())
    
    # Select spots to group (based on the column pseudo_by)
    pseudo_groups = unique(metadata.filt$pseudo_by)
    pseudo_counts = data.frame(row.names = dat$V1)
    
    for (pseudo_unit in pseudo_groups){
      #pseudo_unit = pseudo_groups[1] # debug
      metadata_subgroup = metadata.filt %>% filter(pseudo_by == pseudo_unit)
      subset_idx = which(colnames(dat) %in% metadata_subgroup$barcode)
      if (length(subset_idx)>=nspots){
        cat("\n",paste("Sample:",sample,"Group by:",pseudo_unit,"NumSpots:",length(subset_idx)))
        dat_subgroup = dat[, subset_idx, drop=F]
        counts_subgroup = data.frame(id = rowSums(dat_subgroup, na.rm = T)) ### SUM over spots
        colnames(counts_subgroup) <- paste0(pseudo_unit,"_",length(subset_idx))
        pseudo_counts <- cbind(pseudo_counts, counts_subgroup) 
      }
    }
    return(pseudo_counts)
  }
  pseudo_counts_list <- purrr::map(sample_ids, read_by_donor)
  pseudo_counts = purrr::reduce(pseudo_counts_list, bind_cols) 
  save(pseudo_counts_list,pseudo_counts, file = outputRdata)
}

make_pseudo_matrices_limma_residuals <- function(metadata.filt, 
                                 sample_ids,
                                 outputRdata = paste0(work_dir,"pseudo_bulk/by_donor_section/",output_prefix,"_pseudo_counts.RData"),
                                 nspots = 50){
  # This function will first regress layer in each section matrix
  # Then it will apply stratification and pseudo-bulk summing the residuals 
  library(limma)
  
  read_by_donor <- function(sample){
    # Read each counts file (1 file for each donor)
    #sample = "B16002"
    cnt_file = data.df[data.df$sample_id == sample,]$file
    suppressWarnings(dat <- data.table::fread(cnt_file) %>% as.data.frame())
    # Counts table for the donor
    pseudo_counts = data.frame(row.names = dat$V1)
    # Metadata for the donor
    metadata_donor = unique(metadata.filt[metadata.filt$Bnum == sample,c("barcode","section","annotated_clusters")])
    
    for (section in unique(metadata_donor$section)){
      # Loop over sections of the selected donor
      #section = unique(metadata_donor$section)[1]
      # Create section specific metadata and counts table
      metadata_section = unique(metadata_donor[metadata_donor$section == section,])
      subset_section_idx = which(colnames(dat) %in% metadata_section$barcode)
      dat_section = dat[, subset_section_idx, drop=F]
      # Regress layer from the section counts table
      residuals_matrix <- removeBatchEffect(dat_section, batch = metadata_section$annotated_clusters)
      
      # Select spots to group (based on the column pseudo_by)
      pseudo_groups = unique(metadata.filt$pseudo_by)
      for (pseudo_unit in pseudo_groups){
        # Loop over stratification labels
        #pseudo_unit = pseudo_groups[1] # debug
        # Create strat metadata and residual counts table
        metadata_subgroup = metadata.filt %>% filter(pseudo_by == pseudo_unit)
        subset_idx = which(colnames(residuals_matrix) %in% metadata_subgroup$barcode)
        if (length(subset_idx)>=nspots){ # Cutoff for grouping spots
          cat("\n",paste("Sample:",sample,"Group by:",pseudo_unit,"NumSpots:",length(subset_idx)))
          dat_subgroup = residuals_matrix[, subset_idx, drop=F]
          # SUM over spots
          counts_subgroup = data.frame(id = rowSums(dat_subgroup, na.rm = T)) 
          # Make column name, keeping number of spots used in the grouping
          colnames(counts_subgroup) <- paste0(pseudo_unit,"_",length(subset_idx))
          # Add to the final table
          pseudo_counts <- cbind(pseudo_counts, counts_subgroup) 
        }
      }
    }
    # Return donor specific pseudo-bulk table
    return(pseudo_counts)
  }
  # Get a list of pseudo-bulk tables (1 for each donor)
  pseudo_counts_list <- purrr::map(sample_ids, read_by_donor) 
  # Merge tables 
  pseudo_counts = purrr::reduce(pseudo_counts_list, bind_cols) 
  # Save results
  save(pseudo_counts_list,pseudo_counts, file = outputRdata)
}

make_pseudo_matrices_limma_residuals2 <- function(metadata.filt, 
                                 sample_ids,
                                 outputRdata = paste0(work_dir,"pseudo_bulk/by_donor_section/",output_prefix,"_pseudo_counts.RData"),
                                 nspots = 50){
  # This function will first regress layer in each section matrix
  # Then it will apply stratification and pseudo-bulk summing the residuals 
  library(limma)
  
  read_by_donor <- function(sample){
    # Read each counts file (1 file for each donor)
    #sample = "B16006"
    cnt_file = data.df[data.df$sample_id == sample,]$file
    suppressWarnings(dat <- data.table::fread(cnt_file) %>% as.data.frame())
    # Counts table for the donor
    pseudo_counts = data.frame(row.names = dat$V1)
    # Metadata for the donor
    metadata_donor = unique(metadata.filt[metadata.filt$Bnum == sample,c("barcode","section","annotated_clusters",
                                                                         "log_amyloid_top","log_GFAP_top","log_IBA1_top",
                                                                         "log_amyloid_bottom","log_GFAP_bottom","log_IBA1_bottom",
                                                                         "log_amyloid_avg","log_GFAP_avg","log_IBA1_avg")])
    
    for (section in unique(metadata_donor$section)){
      # Loop over sections of the selected donor
      #section = unique(metadata_donor$section)[1]
      # Create section specific metadata and counts table
      metadata_section = unique(metadata_donor[metadata_donor$section == section,])
      subset_section_idx = which(colnames(dat) %in% metadata_section$barcode)
      dat_section = dat[, subset_section_idx, drop=F]
      variables2keep = model.matrix.lm(~ log_amyloid_avg + log_GFAP_avg + log_IBA1_avg, metadata_section, na.action="na.pass")
      variables2keep = variables2keep[ , colSums(is.na(variables2keep))==0]
      # Regress layer from the section counts table
      residuals_matrix <- removeBatchEffect(dat_section, batch = metadata_section$annotated_clusters, design = variables2keep)
      
      # Select spots to group (based on the column pseudo_by)
      pseudo_groups = unique(metadata.filt$pseudo_by)
      for (pseudo_unit in pseudo_groups){
        # Loop over stratification labels
        #pseudo_unit = pseudo_groups[1] # debug
        # Create strat metadata and residual counts table
        metadata_subgroup = metadata.filt %>% filter(pseudo_by == pseudo_unit)
        subset_idx = which(colnames(residuals_matrix) %in% metadata_subgroup$barcode)
        if (length(subset_idx)>=nspots){ # Cutoff for grouping spots
          cat("\n",paste("Sample:",sample,"Group by:",pseudo_unit,"NumSpots:",length(subset_idx)))
          dat_subgroup = residuals_matrix[, subset_idx, drop=F]
          # SUM over spots
          counts_subgroup = data.frame(id = rowSums(dat_subgroup, na.rm = T)) 
          # Make column name, keeping number of spots used in the grouping
          colnames(counts_subgroup) <- paste0(pseudo_unit,"_",length(subset_idx))
          # Add to the final table
          pseudo_counts <- cbind(pseudo_counts, counts_subgroup) 
        }
      }
    }
    # Return donor specific pseudo-bulk table
    return(pseudo_counts)
  }
  # Get a list of pseudo-bulk tables (1 for each donor)
  pseudo_counts_list <- purrr::map(sample_ids[1], read_by_donor) 
  pseudo_counts_list <- purrr::map(sample_ids, read_by_donor) 
  # Merge tables 
  pseudo_counts = purrr::reduce(pseudo_counts_list, bind_cols) 
  # Save results
  save(pseudo_counts_list,pseudo_counts, file = outputRdata)
}

normalize_pseudo_matrices <- function(metadata.filt, 
                                      inputRdata = paste0(work_dir,"pseudo_bulk/by_donor_section/",output_prefix,"_pseudo_counts.RData"), 
                                      outputRdata = paste0(work_dir,"pseudo_bulk/by_donor_section/",output_prefix,"_","pseudo_counts_norm.RData"),
                                      cpm_cutoff = 1,
                                      cpm_sampleFrac = 0.8){
  ## Perform normalization over pseudo-bulk matrices
  load(inputRdata) # Load "pseudo_counts" 
  library(edgeR)
  # Create metadata at donor level
  metadata_pseudo = unique(metadata.filt %>% dplyr::select(Bnum, RIN, age, ad_reagan, cogdx, diagnosis, cognition, PMI, amyloid, sample_id, librarybatch))
  rownames(metadata_pseudo) = NULL
  
  counts = pseudo_counts
  
  # Filter genes by CPM
  counts_cpm = as.data.frame(cpm(counts))
  keep.exp <- rowSums(counts_cpm > 1) >= (cpm_sampleFrac * ncol(counts_cpm) )
  counts_cpm_filt = counts_cpm[keep.exp,,drop=F]
  counts_filt = counts[keep.exp,,drop=F] #14k genes
  
  # TMM-voom  
  norm <- calcNormFactors(counts_filt, method = "TMM") 
  dge = DGEList(counts_filt, norm.factors = norm)
  counts_voom <- voom(dge)
  tmm_voom <- as.data.frame(counts_voom$E)[,,drop=F]
    
  pseudo_counts_norm = list()
  pseudo_counts_norm$raw_counts = counts
  pseudo_counts_norm$counts = counts_filt
  pseudo_counts_norm$cpm = counts_cpm_filt
  pseudo_counts_norm$tmm_voom = tmm_voom
  
  save(metadata_pseudo, pseudo_counts_norm, file = outputRdata)
}

normalize_residual_pseudo_matrices <- function(metadata.filt, 
                                      inputRdata = paste0(work_dir,"pseudo_bulk/by_donor_section/",output_prefix,"_pseudo_counts.RData"), 
                                      outputRdata = paste0(work_dir,"pseudo_bulk/by_donor_section/",output_prefix,"_","pseudo_counts_norm.RData"),
                                      cpm_cutoff = 1,
                                      cpm_sampleFrac = 0.8){
  ## Perform normalization over pseudo-bulk matrices
  # The difference between this function and "normalize_pseudo_matrices" is that pseudo-bulk tables with counts < 0 are converted to 0
  # This is proposed to deal with pseudo-bulk created using residualized counts (after adjusting for some covariate)
  
  load(inputRdata) # Load "pseudo_counts" 
  library(edgeR)
  # Create metadata at donor level
  metadata_pseudo = unique(metadata.filt %>% dplyr::select(Bnum, RIN, age, ad_reagan, cogdx, diagnosis, cognition, PMI, amyloid, sample_id, librarybatch))
  rownames(metadata_pseudo) = NULL
  
  counts = pseudo_counts
  
  # Filter genes by CPM
  counts_cpm = as.data.frame(cpm(counts))
  keep.exp <- rowSums(counts_cpm > cpm_cutoff) >= (cpm_sampleFrac * ncol(counts_cpm) )
  counts_cpm_filt = counts_cpm[keep.exp,,drop=F]
  counts_filt = counts[keep.exp,,drop=F] #14k genes
  # Convert any negative value to 0
  counts_filt[counts_filt<0] = 0
  
  # TMM-voom  
  norm <- calcNormFactors(counts_filt, method = "TMM") 
  dge = DGEList(counts_filt, norm.factors = norm)
  counts_voom <- voom(dge)
  tmm_voom <- as.data.frame(counts_voom$E)[,,drop=F]
    
  pseudo_counts_norm = list()
  pseudo_counts_norm$raw_counts = counts
  pseudo_counts_norm$counts = counts_filt
  pseudo_counts_norm$cpm = counts_cpm_filt
  pseudo_counts_norm$tmm_voom = tmm_voom
  
  save(metadata_pseudo, pseudo_counts_norm, file = outputRdata)
}

plot_exp_density_per_sample <- function(exp_matrix, title = "", subtitle = "", x_label = "value", color = "#4DBBD5FF"){
  library(ggpubr)
  library(viridis)
  #exp_matrix <- cell_type$RPKM.cqn.log2
  exp_matrix <- as.matrix(exp_matrix)
  nsamples <- ncol(exp_matrix)
  colfunc <- grDevices::colorRampPalette(c(color, alpha(color, 0.5)))
  col = alpha(colfunc(nsamples), alpha = 0.1)
  per_sample = reshape2::melt(exp_matrix)
  density_p <- ggplot(per_sample, aes(x=value, color=as.factor(Var2))) + 
    geom_density(show.legend = FALSE) + 
    scale_fill_viridis_d() +
    scale_color_manual(values=col) +
    labs(x = x_label, y = "Density", title = title, subtitle = subtitle) +
    theme_classic()
  return(density_p)
}
```

## Set parameters

```{r}
work_dir = "/pastel/Github_scripts/AD_Spatial_Transcriptome"
data_folder = paste0(work_dir,"/data/alldata")
metadata_file = paste0(work_dir,"/data/metadata.csv")
strat_file = paste0(work_dir,"/data/strat.xlsx")

path_to_save = paste0(work_dir,"/data/pseudo_bulk/")
system(paste0("mkdir -p ", path_to_save))

print(paste0("Results save to: ", path_to_save))
```

## Read files

```{r}
library(data.table)

# Data files are the counts (or normalized counts) files for each individual. 
data_files = list.files(path = data_folder, pattern = "raw*", full.names = T)

# Match the metadata with the count tables 
sample_id = gsub("(.*)_(.*)\\.csv","\\2",data_files)
data.df = data.frame(sample_id = sample_id, file = data_files)
```

Number of spots by layer/brain region

```{r}
metadata = fread(metadata_file, sep = ",") %>% as.data.frame()
table(distinct(metadata[,c("V1","annotated_clusters")])$annotated_clusters, useNA = "ifany")
table(distinct(metadata[,c("V1","brainRegion")])$brainRegion, useNA = "ifany")
```

After merging with stratification tables

```{r}
if(endsWith(strat_file,"xlsx")){
  df_strat = readxl::read_xlsx(strat_file)
}else{
  df_strat = fread(strat_file, sep = ",") %>% as.data.frame()
}

df_strat_m = df_strat %>% 
  pivot_longer(cols = everything(), names_to = "strat", values_to = "barcode") %>% distinct()

metadata_with_strat = metadata %>% inner_join(df_strat_m, by = c("cells"="barcode"))

table(distinct(metadata_with_strat[,c("V1","sample_section","strat")])[,c("sample_section","strat")], useNA = "ifany")
```

## Make Pseudo-bulk matrices 

Define here how to group spots into a pseudocount matrix. For example, if it just by donor, set `metadata$pseudo_by = metadata$projID`, if it's by donor AND region, set `metadata$pseudo_by = paste(metadata$projID,metadata$annotated_clusters, sep = "_")`. 

Normal pseudo-bulk using SUM of stratified spots (50 spots min)

```{r eval=T}
pseudo_by_list = list(strats_50spots_all = c("sample_section", "strat"),
                      strats_50spots_byRegion = c("sample_section", "strat", "brainRegion"))
metadata = metadata_with_strat
for(i in 1:length(pseudo_by_list)){
  # i=1
  metadata.filt = metadata
  output_prefix = names(pseudo_by_list)[i]
  # Remove if empty or NA
  toRemNA = rowSums(is.na(metadata.filt[,pseudo_by_list[[i]],drop=F ]), na.rm = T)>0
  toRemEmpty = rowSums(metadata.filt[,pseudo_by_list[[i]],drop=F ]=="", na.rm = T)>0
  metadata.filt = metadata.filt[!(toRemNA | toRemEmpty),]
  # Make pseudo_by column 
  metadata.filt$pseudo_by <- apply( metadata.filt[,pseudo_by_list[[i]],drop=F ] , 1 , paste , collapse = "_" )
  #length(unique(metadata.filt$pseudo_by))

  to_keep = names(table(metadata.filt$pseudo_by))[table(metadata.filt$pseudo_by)>=50] # useful when generating pseudo-bulk with spots as "samples"
  metadata.filt = metadata.filt[metadata.filt$pseudo_by %in% to_keep,]
  # Check the number of spots per donor/section 
  #table(metadata.filt[,pseudo_by_list[[i]],drop=F])
  # as.data.frame(table(metadata.filt[,pseudo_by_list[[i]],drop=F])) %>% filter(Freq > 0)
  
  # We will read each sample' counts table, and create pseudo-bulk tables for each desired group (e.g. donor).
  make_pseudo_matrices(metadata.filt, data.df$sample_id, outputRdata = paste0(path_to_save,output_prefix,"_pseudo_counts.RData"), nspots = 1)
  
  # Create normalized matrices
  normalize_pseudo_matrices(metadata.filt = metadata.filt, 
                            inputRdata = paste0(path_to_save,output_prefix,"_pseudo_counts.RData"),
                            outputRdata = paste0(path_to_save,output_prefix,"_pseudo_counts_norm.RData"))
  
  ## Make tables
  load(file = paste0(path_to_save,output_prefix,"_pseudo_counts_norm.RData"))
  table_norm <- as.data.frame(pseudo_counts_norm[["tmm_voom"]]) %>% rownames_to_column("gene")
  data.table::fwrite(x = table_norm, sep = "\t", row.names = F, col.names = T, quote = F,
                     file = paste0(path_to_save,output_prefix,".tmm_voom.tsv.gz"))
  table_norm.cpm <- as.data.frame(pseudo_counts_norm[["cpm"]]) %>% rownames_to_column("gene")
  data.table::fwrite(x = table_norm.cpm, sep = "\t", row.names = F, col.names = T, quote = F,
            file = paste0(path_to_save,output_prefix,".cpm.tsv.gz"))
}
```

## Session info

```{r}
sessionInfo()
```
